Quantum Hoare Types

### 2020-02-20 ms-paper/ms-paper.tex
More rearrangement

\section{Running Example}

In Fig. \ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right. It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$. The function first entangles the latter two qubits into a Bell state. Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob. Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice. The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}. The specification that the function must meet is written using a Hoare type, which we describe next.
\section{Hoare Types for Quantum Programming Languages}
In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol. We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.

@@

\begin{figure}[t]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
					\end{minipage}\\
		\textit{Wire Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
		\textit{Propositions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \mathrm{Id_A}(M, N)$
		\end{minipage}\\
%		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
%		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ x = N ~\mathrm{in}~ M $
		\end{minipage}\\
		\textit{Simple Terms} & $a, b$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
		\textit{Circuits} & $\mathcal{C}, \mathcal{D}$ & : & $ \Sigma \to \Sigma' $ \\
		\textit{Proposition Contexts} & $\Delta$ & ::= & $ \cdot \mid P, \Delta$ \\
		\textit{Label Contexts} & $\Sigma$ & ::= & $ \cdot \mid l : \type{Qubit}, \Sigma$ \\
%		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
%			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
%			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
%		\end{minipage}\\
%		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
%		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
%		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Gamma$ \\
		\textit{Values} & $V$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
	$ x \mid l \mid \lambda x.M \mid \mathrm{lift}~M \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~V \mid \mathrm{force}(\mathrm{unbox}~V) \mid $
\end{minipage}\\
	\end{tabular}
	\Description{Syntax of Hoare Type Theory from \cite{nanevski_hoare_2008}}
	\caption{Syntax of Quantum (Linear) Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

@@

@@

>>>>

\section{Hoare Types for Quantum Programming Languages}
% In this section we describe Quantum Hoare Types using an example of programming quantum teleportation protocol. We start with the background necessary to understand teleportation and then present how Quantum Hoare Types help us write and verify the  correctness of the protocol.

@@

@@

\begin{figure}[h]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
			$ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid A \otimes B \mid \Pi x{ : }A.B \mid ~!A  \mid \type{Circ}(S, U) \mid \{P\} x{ : }A \{Q\}$
					\end{minipage}\\
		\textit{Wire Types} & $S, U$ & ::= & $ \type{Unit} \mid \type{Bit} \mid \type{Qubit} \mid S \otimes U$ \\
		\textit{Propositions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid P \supset Q \mid \neg P \mid \exists x{:}A.P \mid \forall x{:}A.P \mid \mathrm{Id_A}(M, N)$
		\end{minipage}\\
%		\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
%		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
		\textit{Terms} & $M, N$ & ::= & \begin{minipage}[t]{0.55\columnwidth}%
			$ c \mid l \mid x \mid \lambda x.M \mid MN \mid \mathrm{unit} \mid (a, \mathcal{C}, b) \mid \mathrm{box_S}~M \mid \mathrm{unbox}~M \mid \mathrm{force}~M \mid \mathrm{lift}~M \mid (M, N) \mid \mathrm{let}~ x = N ~\mathrm{in}~ M $
		\end{minipage}\\
		\textit{Simple Terms} & $a, b$ & ::= & $ l \mid \mathrm{unit} \mid (a, b)$ \\
		\textit{Circuits} & $\mathcal{C}, \mathcal{D}$ & : & $ \Sigma \to \Sigma' $ \\
		\textit{Proposition Contexts} & $\Delta$ & ::= & $ \cdot \mid P, \Delta$ \\
		\textit{Label Contexts} & $\Sigma$ & ::= & $ \cdot \mid l : \type{Qubit}, \Sigma$ \\
%		\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
%			$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
%			\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
%		\end{minipage}\\
%		\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
%		\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
%		\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
		\textit{Contexts} & $\Gamma$ & ::= & $ \cdot \mid x : A, \Gamma \mid l : \type{Qubit}, \Gamma$ \\
		\textit{Values} & $V$ & ::= & \begin{minipage}[t]{0.5\columnwidth}%
	$ x \mid l \mid \lambda x.M \mid \mathrm{lift}~M \mid (a, \mathcal{C}, b) \mid \mathrm{unbox}~V \mid \mathrm{force}(\mathrm{unbox}~V) \mid $
\end{minipage}\\
	\end{tabular}
	\Description{Syntax of Hoare Type Theory from \cite{nanevski_hoare_2008}}
	\caption{Syntax of Quantum (Linear) Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

@@

In Fig. \ref{fig:teleport}, on the left, we show a function \lstinline{teleport} written in \qwire-like syntax (TODO: add ref) that implements the quantum teleportation circuit shown on the right. It returns one qubit and takes three qubits as arguments: one qubit \lstinline{q} that could be in any unknown state $\ket{\psi}$ which Alice is trying to send to Bob; and two qubits, \lstinline{a} and \lstinline{b}, initially in the state $\ket{0}$. The function first entangles the latter two qubits into a Bell state. Then, Alice performs a Bell measurement of her two qubits \lstinline{q} and \lstinline{a} and sends the resulting classical bits (\lstinline{x} and \lstinline{y}) to Bob. Finally, Bob applies corrective measurement using the quantum gates X and Z controlled by the classical bits that he received from Alice. The result is that Bob's qubit is now in the same state as the initial state of the message qubit \lstinline{q}. The specification that the function must meet is written using a Hoare type, which we described in the previous section.
